\textbf{Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces}

### 2020-03-20 submission/manuscript.tex
References update.

Biofilms formed from microbial epiphytes can contain diverse taxonomic
groups and harbor cell densities from 10\textsuperscript{2} to
10\textsuperscript{7} \si{\cells\per\cm\squared} (Armstrong \emph{et
al.}, 2000; Bengtsson \emph{et al.}, 2010; Burke \emph{et al.}, 2011b).
In such an environment a number of positive and negative interactions
between the macrophyte and colonizing microorganisms have been described
(Egan \emph{et al.}, 2013; Hollants \emph{et al.}, 2013; Tarquinio
\emph{et al.}, 2019). Macrophytes can promote growth of associated
microbes by nutrient exudation, while in return microorganisms may
support macrophyte performance through improved nutrient availability,
phytohormone production and protection form toxic compounds, oxidative
stress, biofouling organisms and pathogens (Egan \emph{et al.}, 2013;
Hollants \emph{et al.}, 2013; Tarquinio \emph{et al.}, 2019). Beside
this positive interactions, macrophytes can negatively impact the
associated microbes such as pathogenic bacteria by producing reactive
oxygen species and secondary metabolites (Egan \emph{et al.}, 2013;
Hollants \emph{et al.}, 2013; Tarquinio \emph{et al.}, 2019).

All these ecological roles are carried out by a taxonomically diverse
community of microorganisms. At higher taxonomic ranks a core set of
macrophyte epiphytic taxa was described consisting mainly of
\emph{Alphaproteobacteria}, \emph{Gammaproteobacteria},
\emph{Desulfobacterota}, \emph{Bacteroidota}, \emph{Cyanobacteria},
\emph{Actinobacteriota}, \emph{Firmicutes}, \emph{Planctomycetota},
\emph{Chloroflexi} and \emph{Verrucomicrobiota} (Crump and Koch, 2008;
Tujula \emph{et al.}, 2010; Lachnit \emph{et al.}, 2011). In contrast,
at lower taxonomic ranks host specific microbial communities were
described (Lachnit \emph{et al.}, 2011; Roth-Schulze \emph{et al.},
2016). Recently, it was shown that even different morphological niches
within the same alga had a higher influence on bacterial community
variation than biogeography or environmental factors (Morrissey \emph{et
al.}, 2019). While there is high community variation between host
species it was observed that the majority of metagenome determined
functions were conserved both between host species and individuals
(Burke \emph{et al.}, 2011a; Roth-Schulze \emph{et al.}, 2016). This
discrepancy between taxonomic and functional composition could be
explained by the lottery hypothesis. It postulates that an initial
random colonization step is performed from a set of functionally
equivalent taxonomic groups resulting in taxonomically different
epiphytic communities sharing a core set of functional genes (Burke
\emph{et al.}, 2011a; Roth-Schulze \emph{et al.}, 2016). In addition,
some of the variation in the observed data could be attributed to
different techniques used in various studies, such as different
protocols for epiphytic cell detachment and/or DNA isolation, as no
standard protocol to study epiphytic communities was established
(Ugarelli \emph{et al.}, 2019; Korlević \emph{et al.}, submitted).

The majority of studies describing macrophyte epiphytic communities did
not encompass seasonal changes (Crump and Koch, 2008; Lachnit \emph{et
al.}, 2009; Burke \emph{et al.}, 2011b; Roth-Schulze \emph{et al.},
2016; Ugarelli \emph{et al.}, 2019). In addition, if seasonal changes
were taken into account low temporal frequency and/or methodologies that
do not allow high taxonomic resolution were used (Tujula \emph{et al.},
2010; Lachnit \emph{et al.}, 2011; Miranda \emph{et al.}, 2013; Michelou
\emph{et al.}, 2013). In the present study we describe the seasonal
dynamics of bacterial and archaeal communities on the surfaces of the
seagrass \emph{Cymodocea nodosa} and siphonous macroalgae \emph{Caulerpa
cylindracea} determined on a mostly monthly scale. Bacterial and
archaeal epiphytes were sampled in a meadow of \emph{C. nodosa} invaded
by the invasive \emph{C. cylindracea} and in a locality of only \emph{C.
cylindracea} located in the proximity of the meadow. In addition, for
comparison, the community of the ambient seawater was characterized.

@@

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of eukaryotic,
chloroplast, mitochondrial and no relative sequences
(\autoref{supp-nseq_notus}). A total of 35 samples originating from
epiphytic archaeal and bacterial communities associated with surfaces of
the seagrass \emph{C. nodosa} and macroalga \emph{C. cylindracea} were
analyzed. In addition, 18 samples (one of the samples was sequenced two
times) originating from picoplankton archaeal and bacterial communities
in the ambient seawater were also processed for comparison. The number
of reads per sample ranged between 8,409 and 77,465 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off that is a common
observation in high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences mentioned before reads were clustered into 28,691 different
OTUs at a similarity level of 97 \si{\percent}. Read numbers were
normalized to the minimum number of sequences, 8,409
(\autoref{supp-nseq_notus}), through rarefaction resulting in 17,071
different OTUs that ranged from 386 to 2,009 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes of richness
and diversity the Observed Number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,679.5 ±
127.4 OTUs{]} and monospecific {[}Number of OTUs, 1,736.3 ± 156.5
OTUs{]}), middle values for \emph{C. nodosa} (Number of OTUs, 1,059 ±
216.2 OTUs) and lower values for picoplankton communities in the ambient
seawater (Number of OTUs, 528.2 ± 135.4 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not show such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly larger values
in Spring and Summer in comparison to Autumn and Winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient were, respectively, calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from a
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared values were
calculated between seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in the middle between these two
extremes. To assess seasonal changes in the proportion of shared OTUs
and communities the Jaccard's and Bray-Curtis Similarity Coefficients
were calculated between consecutive sampling points (\autoref{shared}).
Both coefficients showed similar trends. Temporal proportional changes
were more pronounced for seawater in comparison to \emph{C. nodosa} and
especially \emph{C. cylindracea} associated communities
(\autoref{shared}). In addition, only 0.4 -- 1.0 \si{\percent} of OTUs
from each surface associated community were found at every time point.
These OTUs also made a high proportion of total sequences (39.9 -- 51.8
\si{\percent}). To further disentangle the environmental and seasonal
community dissimilarity a Principal Coordinates Analysis (PCoA) based on
Bray-Curtis distances and OTU abundances was applied. It showed a clear
separation between planktonic and surface associated communities
(\autoref{pcoa}). In addition, a separation of epiphytic bacterial and
archaeal communities based on host species was determined. This
separation was further supported by ANOSIM (R = 0.96, \emph{p}
\textless{} 0.001). Seasonal changes of \emph{C. nodosa} associated
communities indicated a separation between Spring, Summer and
Autumn/Winter samples (ANOSIM, R = 0.53, \emph{p} \textless{} 0.01). For
\emph{C. cylindracea} associated communities a separation between Summer
and Autumn/Winter/Spring samples was observed that was not so strongly
supported (ANOSIM, R = 0.30, \emph{p} \textless{} 0.05)
(\autoref{pcoa}).

@@

\emph{Cyanobacteria} related sequences were comprising, on average, 5.5
± 4.4 \si{\percent} of total sequences (\autoref{cyano}). Higher
proportions were found for \emph{C. nodosa} (16.4 ± 5.3 \si{\percent})
and \emph{C. cylindracea} (mixed {[} (7.7 ± 3.9 \si{\percent}){]} and
monospecific {[} (7.8 ± 2.4 \si{\percent}){]}) associated communities in
autumn and for seawater communities in winter (8.8 ± 7.4 \si{\percent}).
Large taxonomic differences between surface associated and seawater
cyanobacterial communities were observed. Seawater communities were
mainly comprised of \emph{Cyanobium} and \emph{Synechococcus}, while
surface associated communities were consisted of \emph{Pleurocapsa} and
sequences without known relatives within \emph{Cyanobacteriia}
(\autoref{cyano}). In addition, seasonal changes in surface associated
communities were observed with \emph{Pleurocapsa} and no relative
\emph{Cyanobacteriia} comprising larger proportions in autumn and winter
and \emph{Acrophormium}, \emph{Phormidesmis} and no relative
\emph{Nodosilineaceae} in spring and summer (\autoref{cyano}).

@@

Since the colonization of macrophyte surfaces is performed from a pool
of prokaryotic cells from the ambient seawater, it was interesting to
see to which extent these two communities differ. We observed a strong
differentiation between the surface attached and seawater communities at
the level of OTUs that is in agreement with most published studies
(Burke \emph{et al.}, 2011b; Michelou \emph{et al.}, 2013; Roth-Schulze
\emph{et al.}, 2016; Crump \emph{et al.}, 2018; Ugarelli \emph{et al.},
2019). These data indicate that marine macrophytes are selecting, from a
pool of seawater microbial taxa, the one that can colonize and
proliferate on their surfaces (Salaün \emph{et al.}, 2012; Michelou
\emph{et al.}, 2013). In contrast to these findings Fahimipour \emph{et
al.} (2017) found, in a global study of \emph{Zostera marina},
similarities between leaves and seawater samples. Discrepancies between
our data and this study could be explained by differences in studied
seagrass species, methodological variations or biogeographic trends as
Fahimipour \emph{et al.} (2017) were analyzing samples from different
locations throughout the northern hemisphere. It is possible that
ambient seawater and leaves communities from the same location are
differing but are still more similar to each other when compared to
other sampling locations. Indeed, it was found that prokaryotic
communities vary substantially between different sampling sites
(Bengtsson \emph{et al.}, 2017). When the taxonomic composition at high
ranks was analyzed no such strong differentiation was noticed. Phyla and
classes such as: \emph{Actinobacteriota}, \emph{Bacteroidota},
\emph{Cyanobacteria}, \emph{Alphaproteobacteria},
\emph{Gammaproteobacteria} and \emph{Verrucomicrobiota} were described
that is in agreement with previously reported data (Burke \emph{et al.},
2011b; Egan \emph{et al.}, 2013; Michelou \emph{et al.}, 2013). In
contrast, when low taxonomic ranks were analyzed (i.g. family and genus)
a strong differentiation was observed. A similar differentiation at
lower taxonomic ranks was described for other species of macrophytes
(Egan \emph{et al.}, 2013; Michelou \emph{et al.}, 2013; Ugarelli
\emph{et al.}, 2019).

Beside differences between seawater and surface associated communities,
there were discussions if the prokaryotic epiphytic community is
host-specific or there are generalists taxa characteristic to all or
many macrophytes (Egan \emph{et al.}, 2013). Similarly to previously
described differences between seawater and surface attached communities,
at high taxonomic ranks no strong differentiation between communities
associated with different host was observed. The only high rank phylum
that was differing between \emph{C. nodosa} and \emph{C. cylindracea}
was \emph{Desulfobacterota}, whose sequences were more abundant in the
\emph{C. cylindracea} associated community. As already mentioned, the
rhizoids and part of the stolon are in contact with the sediment, so
\emph{Desulfobacterota} are probably a part of the epiphytic community
that was in contact with the sediment. Similar high rank taxa found in
this study were described to be specific for other species of
macrophytes (Burke \emph{et al.}, 2011b; Lachnit \emph{et al.}, 2011;
Bengtsson \emph{et al.}, 2017). In contrast to high taxonomic ranks, a
substantial differentiation between host specific communities was found,
which supports the host-specific hypothesis. Similar host-specificity
was observed for different species of macroalgae and seagrasses (Lachnit
\emph{et al.}, 2011; Roth-Schulze \emph{et al.}, 2016; Ugarelli \emph{et
al.}, 2019; Morrissey \emph{et al.}, 2019). Taken together, at high
taxonomic ranks a core set of taxa could be described that is
characteristic for all or many macrophytes, while at low taxonomic ranks
a community specific to host species could be identified (Egan \emph{et
al.}, 2013).

Seasonal richness changes in the epiphytic community were substantial as
could be observed in the proportion of OTUs that could be found at every
sampling time (\(\leq\) 1.0 \si{\percent}). Interestingly, these OTUs
were accounting for a high proportion of sequences (\(\leq\) 51.8
\si{\percent}). A very similar proportion of persistent OTUs and their
sequence contribution was reported in high frequency studies describing
seasonal picoplankton changes (Gilbert \emph{et al.}, 2009, 2012). In
comparison to the seawater community, a lower degree of seasonal shifts
was observed for the surface associated communities. It seems,
microniches on the surfaces of macrophytes are providing more stable
conditions in comparison to the seawater. At the level of OTUs seasonal
changes of \emph{C. nodosa} and \emph{C. cylindracea} associated
communities were identified that could be linked to the growth cycle of
the seagrass and macroalgae (M. Najdek, personal communication).
\emph{C. nodosa} was characterized by a Spring community during maximum
seagrass proliferation, a Summer community during a biomass maximum and
a Autumn/Winter community during a biomass senescence. In contrast,
\emph{C. cylindracea} started to proliferate in late Spring and was
characterized only by a Summer community during maximal biomass increase
and by a Autumn/Winter/Spring community when the biomass was at the peak
and the settlement started to subsequently decay. Similar seasonal
changes in the epiphytic community was also described for other
macroalgae (Tujula \emph{et al.}, 2010; Lachnit \emph{et al.}, 2011).
Higher temporal stability of \emph{C. cylindracea} surface communities
in comparison to \emph{C. nodosa} were also observed in the higher
proportion of shared communities between two consecutive sampling
points.

@@

complex carbon sources (Burke \emph{et al.}, 2011b; McIlroy and Nielsen,
2014; Crump \emph{et al.}, 2018). On the other hand, families
\emph{Flavobacteriaceae} and \emph{Rhodothermaceae} showed seasonal
patterns, with \emph{Flavobacteriaceae} being more pronounced from
November to June and \emph{Rhodothermaceae} from June to October. Within
\emph{Alphaproteobacteria} the family \emph{Rhodobacteraceae} was

@@

surfaces and usually is one of the most abundant groups (Burke \emph{et
al.}, 2011b; Michelou \emph{et al.}, 2013; Pujalte \emph{et al.}, 2014).
In addition, \emph{Hyphomonadaceae} were found in all samples.
Interestingly, some of the species within this group contain stalks on
their cells which can be used to attach to the macrophyte surface
(Weidner \emph{et al.}, 2000; Abraham and Rohde, 2014). Within the
\emph{Gammaproteobacteria}, sequences without known representatives were
the most pronounced group present throughout the year. In addition,
\emph{Granulosicoccus} was also found in almost all samples.
\emph{Gammaproteobacteria} are often a major constituent of macrophyte
epiphytic communities (Burke \emph{et al.}, 2011b; Michelou \emph{et
al.}, 2013; Crump \emph{et al.}, 2018). Beside these two groups other
less pronounced taxa showed seasonal and host-specific patterns. For
example, \emph{C. cylindracea} was characterized by \emph{Thioploca}, a
known sulfur sediment bacteria and \emph{Cellvibrionaceae}, a family
whose cultured members are known polysaccharide degraders (Jørgensen and
Gallardo, 1999; Xie \emph{et al.}, 2017). \emph{Desulfobacterota} were
found only associated with \emph{C. cylindracea} and no group within
this phylum showed seasonal patterns. The presence of this phylum only
on \emph{C. cylindracea} is to be expected as part of the epiphytic
community is directly in contact with the sediment. The
\emph{Desulfobacterota} community was dominated by \emph{Desulfatitalea}
and no relative \emph{Desulfocapsaceae}, known sulfate sediment groups
(Kuever, 2014; Higashioka \emph{et al.}, 2015).

In temperate zones marine macrophytes are exhibiting growth cycles, so
it is not surprising that the associated epiphytic microbial community
is undergoing partial seasonal changes. In the present study, we could,
in every analyzed high rank taxa, identify phylogenetic groups that were
present throughout the year and that were comprising most of the
sequences and lower proportion taxa showing seasonal patterns connected
to the macrophyte growth cycle. Studies focusing on functional
comparisons between communities associated with different hosts showed
that the majority of functions could be found in every community,
indicating functional redundancy (Roth-Schulze \emph{et al.}, 2016).
This difference between taxonomic and functional discrepancy was
explained by the lottery hypothesis that hypothesizes an initial random
colonization step performed from a set of functionally equivalent
taxonomic groups (Burke \emph{et al.}, 2011a; Roth-Schulze \emph{et
al.}, 2016). It is possible that functional redundancy is a
characteristic of high abundance taxa detected to be present throughout
the year, while seasonal and/or host-specific functions are an attribute
of taxa displaying successional patterns. Further studies connecting
taxonomy with functional properties will be required to elucidate the
degree of functional redundancy or specificity in epiphytic microbial
communities.

>>>>

Biofilms formed from microbial epiphytes can contain diverse taxonomic
groups and harbor cell densities from 10\textsuperscript{2} to
10\textsuperscript{7} \si{\cells\per\cm\squared} (Armstrong \emph{et
al.}, 2000; Bengtsson \emph{et al.}, 2010; Burke and Thomas \emph{et
al.}, 2011). In such an environment a number of positive and negative
interactions between the macrophyte and colonizing microorganisms have
been described (Egan \emph{et al.}, 2013; Hollants \emph{et al.}, 2013;
Tarquinio \emph{et al.}, 2019). Macrophytes can promote growth of
associated microbes by nutrient exudation, while in return
microorganisms may support macrophyte performance through improved
nutrient availability, phytohormone production and protection form toxic
compounds, oxidative stress, biofouling organisms and pathogens (Egan
\emph{et al.}, 2013; Hollants \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019). Beside this positive interactions, macrophytes can
negatively impact the associated microbes such as pathogenic bacteria by
producing reactive oxygen species and secondary metabolites (Egan
\emph{et al.}, 2013; Hollants \emph{et al.}, 2013; Tarquinio \emph{et
al.}, 2019).

All these ecological roles are carried out by a taxonomically diverse
community of microorganisms. At higher taxonomic ranks a core set of
macrophyte epiphytic taxa was described consisting mainly of
\emph{Alphaproteobacteria}, \emph{Gammaproteobacteria},
\emph{Desulfobacterota}, \emph{Bacteroidota}, \emph{Cyanobacteria},
\emph{Actinobacteriota}, \emph{Firmicutes}, \emph{Planctomycetota},
\emph{Chloroflexi} and \emph{Verrucomicrobiota} (Crump and Koch, 2008;
Tujula \emph{et al.}, 2010; Lachnit \emph{et al.}, 2011). In contrast,
at lower taxonomic ranks host specific microbial communities were
described (Lachnit \emph{et al.}, 2011; Roth-Schulze \emph{et al.},
2016). Recently, it was shown that even different morphological niches
within the same alga had a higher influence on bacterial community
variation than biogeography or environmental factors (Morrissey \emph{et
al.}, 2019). While there is high community variation between host
species it was observed that the majority of metagenome determined
functions were conserved both between host species and individuals
(Burke and Peter Steinberg \emph{et al.}, 2011; Roth-Schulze \emph{et
al.}, 2016). This discrepancy between taxonomic and functional
composition could be explained by the lottery hypothesis. It postulates
that an initial random colonization step is performed from a set of
functionally equivalent taxonomic groups resulting in taxonomically
different epiphytic communities sharing a core set of functional genes
(Burke and Peter Steinberg \emph{et al.}, 2011; Roth-Schulze \emph{et
al.}, 2016). In addition, some of the variation in the observed data
could be attributed to different techniques used in various studies,
such as different protocols for epiphytic cell detachment and/or DNA
isolation, as no standard protocol to study epiphytic communities was
established (Ugarelli \emph{et al.}, 2019; Korlević \emph{et al.},
submitted).

The majority of studies describing macrophyte epiphytic communities did
not encompass seasonal changes (Crump and Koch, 2008; Lachnit \emph{et
al.}, 2009; Burke and Thomas \emph{et al.}, 2011; Roth-Schulze \emph{et
al.}, 2016; Ugarelli \emph{et al.}, 2019). In addition, if seasonal
changes were taken into account low temporal frequency and/or
methodologies that do not allow high taxonomic resolution were used
(Tujula \emph{et al.}, 2010; Lachnit \emph{et al.}, 2011; Miranda
\emph{et al.}, 2013; Michelou \emph{et al.}, 2013). In the present study
we describe the seasonal dynamics of bacterial and archaeal communities
on the surfaces of the seagrass \emph{Cymodocea nodosa} and siphonous
macroalgae \emph{Caulerpa cylindracea} determined on a mostly monthly
scale. Bacterial and archaeal epiphytes were sampled in a meadow of
\emph{C. nodosa} invaded by the invasive \emph{C. cylindracea} and in a
locality of only \emph{C. cylindracea} located in the proximity of the
meadow. In addition, for comparison, the community of the ambient
seawater was characterized.

@@

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of eukaryotic,
chloroplast, mitochondrial and no relative sequences
(\autoref{supp-nseq_notus}). A total of 35 samples originating from
epiphytic archaeal and bacterial communities associated with surfaces of
the seagrass \emph{C. nodosa} and macroalga \emph{C. cylindracea} were
analyzed. In addition, 18 samples (one of the samples was sequenced two
times) originating from picoplankton archaeal and bacterial communities
in the ambient seawater were also processed for comparison. The number
of reads per sample ranged between 8,409 and 77,463 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off that is a common
observation in high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences mentioned before reads were clustered into 28,702 different
OTUs at a similarity level of 97 \si{\percent}. Read numbers were
normalized to the minimum number of sequences, 8,409
(\autoref{supp-nseq_notus}), through rarefaction resulting in 17,172
different OTUs that ranged from 366 to 2,043 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes of richness
and diversity the Observed Number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,683.2 ±
130.9 OTUs{]} and monospecific {[}Number of OTUs, 1,737.2 ± 166.4
OTUs{]}), middle values for \emph{C. nodosa} (Number of OTUs, 1,058.4 ±
216.8 OTUs) and lower values for picoplankton communities in the ambient
seawater (Number of OTUs, 529.5 ± 143.0 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not show such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly larger values
in Spring and Summer in comparison to Autumn and Winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient were, respectively, calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from a
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.78), while lower shared values were
calculated between seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in the middle between these two
extremes. To assess seasonal changes in the proportion of shared OTUs
and communities the Jaccard's and Bray-Curtis Similarity Coefficients
were calculated between consecutive sampling points (\autoref{shared}).
Both coefficients showed similar trends. Temporal proportional changes
were more pronounced for seawater in comparison to \emph{C. nodosa} and
especially \emph{C. cylindracea} associated communities
(\autoref{shared}). In addition, only 0.4 -- 1.0 \si{\percent} of OTUs
from each surface associated community were found at every time point.
These OTUs also made a high proportion of total sequences (40.8 -- 52.2
\si{\percent}). To further disentangle the environmental and seasonal
community dissimilarity a Principal Coordinates Analysis (PCoA) based on
Bray-Curtis distances and OTU abundances was applied. It showed a clear
separation between planktonic and surface associated communities
(\autoref{pcoa}). In addition, a separation of epiphytic bacterial and
archaeal communities based on host species was determined. This
separation was further supported by ANOSIM (R = 0.96, \emph{p}
\textless{} 0.001). Seasonal changes of \emph{C. nodosa} associated
communities indicated a separation between Spring, Summer and
Autumn/Winter samples (ANOSIM, R = 0.53, \emph{p} \textless{} 0.01). For
\emph{C. cylindracea} associated communities a separation between Summer
and Autumn/Winter/Spring samples was observed that was not so strongly
supported (ANOSIM, R = 0.30, \emph{p} \textless{} 0.01)
(\autoref{pcoa}).

@@

\emph{Cyanobacteria} related sequences were comprising, on average, 5.5
± 4.4 \si{\percent} of total sequences (\autoref{cyano}). Higher
proportions were found for \emph{C. nodosa} (16.4 ± 5.3 \si{\percent})
and \emph{C. cylindracea} (mixed {[} (7.7 ± 3.9 \si{\percent}){]} and
monospecific {[} (7.8 ± 2.4 \si{\percent}){]}) associated communities in
autumn and for seawater communities in winter (8.8 ± 7.5 \si{\percent}).
Large taxonomic differences between surface associated and seawater
cyanobacterial communities were observed. Seawater communities were
mainly comprised of \emph{Cyanobium} and \emph{Synechococcus}, while
surface associated communities were consisted of \emph{Pleurocapsa} and
sequences without known relatives within \emph{Cyanobacteriia}
(\autoref{cyano}). In addition, seasonal changes in surface associated
communities were observed with \emph{Pleurocapsa} and no relative
\emph{Cyanobacteriia} comprising larger proportions in autumn and winter
and \emph{Acrophormium}, \emph{Phormidesmis} and no relative
\emph{Nodosilineaceae} in spring and summer (\autoref{cyano}).

@@

Since the colonization of macrophyte surfaces is performed from a pool
of prokaryotic cells from the ambient seawater, it was interesting to
see to which extent these two communities differ. We observed a strong
differentiation between the surface attached and seawater communities at
the level of OTUs that is in agreement with most published studies
(Burke and Thomas \emph{et al.}, 2011; Michelou \emph{et al.}, 2013;
Roth-Schulze \emph{et al.}, 2016; Crump \emph{et al.}, 2018; Ugarelli
\emph{et al.}, 2019). These data indicate that marine macrophytes are
selecting, from a pool of seawater microbial taxa, the one that can
colonize and proliferate on their surfaces (Salaün \emph{et al.}, 2012;
Michelou \emph{et al.}, 2013). In contrast to these findings Fahimipour
\emph{et al.} (2017) found, in a global study of \emph{Zostera marina},
similarities between leaves and seawater samples. Discrepancies between
our data and this study could be explained by differences in studied
seagrass species, methodological variations or biogeographic trends as
Fahimipour \emph{et al.} (2017) were analyzing samples from different
locations throughout the northern hemisphere. It is possible that
ambient seawater and leaves communities from the same location are
differing but are still more similar to each other when compared to
other sampling locations. Indeed, it was found that prokaryotic
communities vary substantially between different sampling sites
(Bengtsson \emph{et al.}, 2017). When the taxonomic composition at high
ranks was analyzed no such strong differentiation was noticed. Phyla and
classes such as: \emph{Actinobacteriota}, \emph{Bacteroidota},
\emph{Cyanobacteria}, \emph{Alphaproteobacteria},
\emph{Gammaproteobacteria} and \emph{Verrucomicrobiota} were described
that is in agreement with previously reported data (Burke and Thomas
\emph{et al.}, 2011; Egan \emph{et al.}, 2013; Michelou \emph{et al.},
2013). In contrast, when low taxonomic ranks were analyzed (i.g. family
and genus) a strong differentiation was observed. A similar
differentiation at lower taxonomic ranks was described for other species
of macrophytes (Egan \emph{et al.}, 2013; Michelou \emph{et al.}, 2013;
Ugarelli \emph{et al.}, 2019).

Beside differences between seawater and surface associated communities,
there were discussions if the prokaryotic epiphytic community is
host-specific or there are generalists taxa characteristic to all or
many macrophytes (Egan \emph{et al.}, 2013). Similarly to previously
described differences between seawater and surface attached communities,
at high taxonomic ranks no strong differentiation between communities
associated with different host was observed. The only high rank phylum
that was differing between \emph{C. nodosa} and \emph{C. cylindracea}
was \emph{Desulfobacterota}, whose sequences were more abundant in the
\emph{C. cylindracea} associated community. As already mentioned, the
rhizoids and part of the stolon are in contact with the sediment, so
\emph{Desulfobacterota} are probably a part of the epiphytic community
that was in contact with the sediment. Similar high rank taxa found in
this study were described to be specific for other species of
macrophytes (Burke and Thomas \emph{et al.}, 2011; Lachnit \emph{et
al.}, 2011; Bengtsson \emph{et al.}, 2017). In contrast to high
taxonomic ranks, a substantial differentiation between host specific
communities was found, which supports the host-specific hypothesis.
Similar host-specificity was observed for different species of
macroalgae and seagrasses (Lachnit \emph{et al.}, 2011; Roth-Schulze
\emph{et al.}, 2016; Ugarelli \emph{et al.}, 2019; Morrissey \emph{et
al.}, 2019). Taken together, at high taxonomic ranks a core set of taxa
could be described that is characteristic for all or many macrophytes,
while at low taxonomic ranks a community specific to host species could
be identified (Egan \emph{et al.}, 2013).

Seasonal richness changes in the epiphytic community were substantial as
could be observed in the proportion of OTUs that could be found at every
sampling time (\(\leq\) 1.0 \si{\percent}). Interestingly, these OTUs
were accounting for a high proportion of sequences (\(\leq\) 52.2
\si{\percent}). A very similar proportion of persistent OTUs and their
sequence contribution was reported in high frequency studies describing
seasonal picoplankton changes (Gilbert \emph{et al.}, 2009, 2012). In
comparison to the seawater community, a lower degree of seasonal shifts
was observed for the surface associated communities. It seems,
microniches on the surfaces of macrophytes are providing more stable
conditions in comparison to the seawater. At the level of OTUs seasonal
changes of \emph{C. nodosa} and \emph{C. cylindracea} associated
communities were identified that could be linked to the growth cycle of
the seagrass and macroalgae (M. Najdek, personal communication).
\emph{C. nodosa} was characterized by a Spring community during maximum
seagrass proliferation, a Summer community during a biomass maximum and
a Autumn/Winter community during a biomass senescence. In contrast,
\emph{C. cylindracea} started to proliferate in late Spring and was
characterized only by a Summer community during maximal biomass increase
and by a Autumn/Winter/Spring community when the biomass was at the peak
and the settlement started to subsequently decay. Similar seasonal
changes in the epiphytic community was also described for other
macroalgae (Tujula \emph{et al.}, 2010; Lachnit \emph{et al.}, 2011).
Higher temporal stability of \emph{C. cylindracea} surface communities
in comparison to \emph{C. nodosa} were also observed in the higher
proportion of shared communities between two consecutive sampling
points.

@@

complex carbon sources (Burke and Thomas \emph{et al.}, 2011; McIlroy
and Nielsen, 2014; Crump \emph{et al.}, 2018). On the other hand,
families \emph{Flavobacteriaceae} and \emph{Rhodothermaceae} showed
seasonal patterns, with \emph{Flavobacteriaceae} being more pronounced
from November to June and \emph{Rhodothermaceae} from June to October.
Within \emph{Alphaproteobacteria} the family \emph{Rhodobacteraceae} was

@@

surfaces and usually is one of the most abundant groups (Burke and
Thomas \emph{et al.}, 2011; Michelou \emph{et al.}, 2013; Pujalte
\emph{et al.}, 2014). In addition, \emph{Hyphomonadaceae} were found in
all samples. Interestingly, some of the species within this group
contain stalks on their cells which can be used to attach to the
macrophyte surface (Weidner \emph{et al.}, 2000; Abraham and Rohde,
2014). Within the \emph{Gammaproteobacteria}, sequences without known
representatives were the most pronounced group present throughout the
year. In addition, \emph{Granulosicoccus} was also found in almost all
samples. \emph{Gammaproteobacteria} are often a major constituent of
macrophyte epiphytic communities (Burke and Thomas \emph{et al.}, 2011;
Michelou \emph{et al.}, 2013; Crump \emph{et al.}, 2018). Beside these
two groups other less pronounced taxa showed seasonal and host-specific
patterns. For example, \emph{C. cylindracea} was characterized by
\emph{Thioploca}, a known sulfur sediment bacteria and
\emph{Cellvibrionaceae}, a family whose cultured members are known
polysaccharide degraders (Jørgensen and Gallardo, 1999; Xie \emph{et
al.}, 2017). \emph{Desulfobacterota} were found only associated with
\emph{C. cylindracea} and no group within this phylum showed seasonal
patterns. The presence of this phylum only on \emph{C. cylindracea} is
to be expected as part of the epiphytic community is directly in contact
with the sediment. The \emph{Desulfobacterota} community was dominated
by \emph{Desulfatitalea} and no relative \emph{Desulfocapsaceae}, known
sulfate sediment groups (Kuever, 2014; Higashioka \emph{et al.}, 2015).

In temperate zones marine macrophytes are exhibiting growth cycles, so
it is not surprising that the associated epiphytic microbial community
is undergoing partial seasonal changes. In the present study, we could,
in every analyzed high rank taxa, identify phylogenetic groups that were
present throughout the year and that were comprising most of the
sequences and lower proportion taxa showing seasonal patterns connected
to the macrophyte growth cycle. Studies focusing on functional
comparisons between communities associated with different hosts showed
that the majority of functions could be found in every community,
indicating functional redundancy (Roth-Schulze \emph{et al.}, 2016).
This difference between taxonomic and functional discrepancy was
explained by the lottery hypothesis that hypothesizes an initial random
colonization step performed from a set of functionally equivalent
taxonomic groups (Burke and Peter Steinberg \emph{et al.}, 2011;
Roth-Schulze \emph{et al.}, 2016). It is possible that functional
redundancy is a characteristic of high abundance taxa detected to be
present throughout the year, while seasonal and/or host-specific
functions are an attribute of taxa displaying successional patterns.
Further studies connecting taxonomy with functional properties will be
required to elucidate the degree of functional redundancy or specificity
in epiphytic microbial communities.
